\newpage
\tableofcontents 
\newpage
\listoftables 
\newpage
\listoffigures 
\newpage

```{r setup, include=FALSE}
# Set your working directory

# Load your packages
library(tidyverse);library(dataRetrieval);library(lubridate);library(remotes)
library(ggplot2);library(ggthemes);library(RColorBrewer)
library(rlang);library(Kendall)
#install_github("USEPA/StreamCatTools", build_vignettes=TRUE, force=TRUE)
library("StreamCatTools") # This imports EPA's tool for StreamCat data

```

``` {r import data}
# Load the data

# Pulling NLCD from StreamCat for every COMID in TX for every year available

years <- c(2001, 2004, 2006, 2008, 2011, 2013, 2016, 2019)
dates <- c('20010101', '20040101', '20060101', '20080101', '20110101', 
           '20130101', '20160101', '20190101') # Best to have ymd format
length(years) # 8 years

nlcd <- data.frame() # Create empty dataframe to be filled

for (i in 1:8) { # For each year for which there are NLCD...
  nlcd_i <- sc_nlcd(years[i], state = "TX") # Create temporary df for this year
  nlcd_i$Year = dates[i] # Denote which year these data are from
  for (col in 1:ncol(nlcd_i)){
    colnames(nlcd_i)[col] <- sub(years[i], "", colnames(nlcd_i)[col])
  } # All the column names have the year in them. I want them removed
  nlcd <- rbind(nlcd, nlcd_i) # Adjoin to growing dataframe of all NLCD
  rm(nlcd_i) # Remove the temporary dataframe
}

#summary(nlcd) # Check the resulting dataframe
# All the fields starting with "PCT" should have a max of 100.00
# Year should run from min: 2001 to max: 2019
# I note there are a lot of NA's; we'll remove those later

# List of Eurycea species found in each COMID (from my MP)
species <- read.csv("species.csv", stringsAsFactors = TRUE)
#summary(species) # This gives a sense of which species have most presences

# Pull BFI for every COMID in TX (this is comparably easy)
bfi <- sc_get_data(metric = "BFI", state = "TX")
#summary(bfi) # Again we see a lot of NA's, to be removed later

# Assemble the final dataframe
nlcd_bfi <- merge(nlcd, bfi) # Intermediary df with the env. variables
env_vars <- merge(species, nlcd_bfi) # Adjoin to salamander presences
rm(nlcd_bfi)#;rm(bfi);rm(nlcd)

```

# Rationale and Research Questions

Genetic testing has delineated fourteen species of Eurycea cave salamanders in central Texas, three of which are undescribed (Devitt et al. 2019). These undescribed species are new to science and therefore enjoy no protections under the Endangered Species Act (ESA). The first of these species was the subject of a petition by Save our Springs Alliance (SOS) to the U.S. Fish and Wildlife Service (USFWS) for its protection (Datri 2021). This petition passed its initial review and USFWS is now considering the species for protection (FWS 2022). Anthropogenic development including land cover change and groundwater overdraft pose the largest threats to Eurycea species by destroying forests that protect their habitats and draining the groundwater in which they live (Datri 2021).

During the 2022-2023 school year, I worked with Rory Kuczek on a NSOE master's project for SOS to facilitate and strengthen future petitions by analyzing habitat threats to these undescribed species. Kuczek and I used MaxEnt and nonmetric multidimensional scaling (NMS) statistical methods to test the association of Eurycea species presences with 13 catchment-level environmental metrics from the StreamCat dataset. Both methods identified that undescribed Eurycea species 2 and 3 presences were most dependent on base flow index (BFI), a ratio of groundwater discharge to total surface flow.

Building on these results, I intend to test several research questions related to habitat degradation for Eurycea species 2 and 3:

1. Has development increased over time across Texas?

2. Has development increased over time in Eurycea habitats?

3. Has development increased over time in new species' habitats?

4. Is development in Texas associated with base flow index?

Answers to these questions should paint a clearer picture of development in Texas and in Eurycea habitat.

\newpage

# Dataset Information

The StreamCat dataset details over 600 natural and anthropogenic metrics for streams in the contintental US at the catchment and watershed levels. The dataset includes robust web services such as a REST API and R package to facilitate data management. StreamCat is maintained by EPA as part of the National Aquatic Resource Surveys. Use of this dataset for analysis of Eurycea salamander species was recommended by John Fay.

The StreamCat package and API provide metrics listed by COMID at the catchment and watersheld level, along with catchment and watershed area. The sc_nlcd function returns all NLCD metrics, in this format, for a given year and area of interest. Column names include the metrics' year of acquisition, which made it difficult to join the data into one dataframe and graph metrics over time. 

To format the data as desired, I wrote a for loop which incrementally obtained NLCD from each year available, added a new "Year" column to record the year, altered the column names to remove the year, and then joined the resulting data to a growing dataframe of NLCD by year. The result is a large dataframe with all NLCD metrics for the specified area of interest (Texas) and all years available.

I subsequently wrangled this dataframe to remove duplicates and NA values, change the "Year" column into a date object, and select only the relevant columns: COMID, Species, Year, BFI (at the catchment level), and summed percents of developed and forested land cover.

Salamander presences were obtained from scientific literature (Devitt et al. 2019) and expert communication in government petitions (Datri 2021).

```{r data wrangling}

Eurycea <-
  env_vars %>%
  distinct() %>%
  drop_na() %>%
  mutate(Year = ymd(Year)) %>%
  mutate(BFI = BFICAT) %>% # We're working at the catchment level only
  mutate(PercentDevelop = PCTURBLOCAT + PCTURBMDCAT + PCTURBHICAT + PCTURBOPCAT) %>%
  mutate(PercentForest = PCTCONIFCAT + PCTDECIDCAT + PCTMXFSTCAT) %>%
  select(COMID, Species, Year, BFI, PercentDevelop, PercentForest)

Eurycea.only <-  # Dataframe of only salamander presences
  Eurycea %>%
  filter(Species != "background")

Eurycea.new <- # Dataframe of only undescribed species
  Eurycea %>%
  filter(Species %in% c("sp1", "sp2", "sp3"))

Dev.yearly <- 
  Eurycea %>%
  group_by(Year) %>%
  summarise(Develop.Mean = mean(PercentDevelop))

Dev.yearly.Eurycea <- # Dataframe of only salamander presences
  Eurycea.only %>%
  group_by(Year, Species) %>%
  summarise(Develop.Mean = mean(PercentDevelop))

Dev.yearly.new <- # Dataframe of only undescribed species
  Eurycea.new %>%
  group_by(Year, Species) %>%
  summarise(Develop.Mean = mean(PercentDevelop))

```

\newpage

# Exploratory Analysis 

```{r development plots, echo=FALSE}

ggplot(Eurycea.only, aes(x = Year, 
                         y = PercentDevelop, 
                         color = Species)) + 
                     geom_point() + 
  # Add line of best fit for each species
 stat_smooth(method = "lm", 
              formula = y ~ x, 
              se = FALSE,
              alpha = 0.5) +
  xlab("Year") +
  ylab("% Developed Land Cover") + 
  ggtitle("Development in Eurycea Habitat over Time") +
  theme_light()

ggplot(Eurycea.new, aes(x = Year, 
                         y = PercentDevelop, 
                         color = Species)) + 
                     geom_point() + 
  # Add line of best fit for each species
 stat_smooth(method = "lm", 
              formula = y ~ x, 
              se = FALSE,
              alpha = 0.5) +
  xlab("Year") +
  ylab("% Developed Land Cover") + 
  ggtitle("Development in Undescribed Eurycea Species Habitats over Time") +
  theme_light()
# I could set ylim to remove outliers from this graph
# But I think those outliers are an important part of this story

```

This first plot shows a clear upward trend in development within catchments from 2001 to 2019 for most Eurycea species. The slope of this trend is more extreme for some species, especially E. neotenes. (I attempted to configure a colorblind-friendly color palette, but symbolizing 14 colors is a tall ask even for broader palettes.)

The second plot shows development over time for the three undescribed species. Initial takeaways are that sp. 3 has the highest percentage of development within its catchments from 2001 to 2019, followed by sp. 2 and sp. 1. It is also clear that known sites for these species have dramatic outliers, with one sp. 3 site routinely scoring at around 30-35% developed land cover, whereas most sites have 10% or less development.

```{r forest plots, echo=FALSE}

ggplot(Eurycea.only, aes(x = Year, 
                         y = PercentForest, 
                         color = Species)) + 
                     geom_point() + 
  # Add line of best fit for each species
 stat_smooth(method = "lm", 
              formula = y ~ x, 
              se = FALSE,
              alpha = 0.5) +
  xlab("Year") +
  ylab("% Forested Land Cover") + 
  ggtitle("Forest in Eurycea Habitat over Time") + 
  theme_light()

ggplot(Eurycea.new, aes(x = Year, 
                         y = PercentForest, 
                         color = Species)) + 
                     geom_point() + 
  # Add line of best fit for each species
 stat_smooth(method = "lm", 
              formula = y ~ x, 
              se = FALSE,
              alpha = 0.5) +
  xlab("Year") +
  ylab("% Forested Land Cover") + 
  ggtitle("Forest in Undescribed Eurycea Species Habitats over Time") +
  theme_light()

```

Conversely, a plot of forested land cover over time shows a clear downward trend for most Eurycea species. This trend is most dramatic, again, for E. neotenes. Together, these plots suggest that most deforestation and development is occurring in E. neotenes catchments relative to other Eurycea species.

The second plot again restricts this analysis to the undescribed species. Forest cover remains relatively constant for sp. 2 (20%) and sp. 3 (40%) over time, but shows a downward trend for sp. 1.

\newpage

# Analysis

I set out to test whether PercentDevelop had changed over time. However, I realized that time-series analysis requires data from continuous, regular time intervals. NLCD is only assessed every 2-3 years, which makes it poorly suited for this task. However, I soldiered on. These results are crude and most likely not fully reliable.

## Question 1: Has development increased over time across Texas?

```{r development change over time}

Dev.yearly.ts <- ts(Dev.yearly$Develop.Mean, 
                    start = c(2001,1,1), end = c(2019,1,1), frequency = 1)
plot(Dev.yearly.ts) # Trend is thrown off by the missing values

Dev.yearly.trend <- Kendall(Dev.yearly$Year, Dev.yearly$Develop.Mean)
summary(Dev.yearly.trend) # p-value = 0.00083661

ggplot(Dev.yearly, aes(x = Year, 
                       y = Develop.Mean)) + 
     stat_smooth(method = "lm", 
                 formula = y ~ x, 
                 se = TRUE,
                 alpha = 0.5) +
     xlab("Year") +
     ylab("Mean Developed Land Cover (%)") + 
     ggtitle("Texas Development in Stream Catchments over Time")

```

A simple Kendall time-series analysis of the data shows a clear upward trend in development throughout Texas stream catchments. The test achieves significance (two-sided p-value = 0.00083661).

## Question 2: Has development increased over time in Eurycea habitats?

```{r Eurycea change over time}

Dev.yearly.Eurycea.ts <- ts(Dev.yearly.Eurycea$Develop.Mean, 
                    start = c(2001,1,1), end = c(2019,1,1), frequency = 1)
plot(Dev.yearly.Eurycea.ts)

Dev.yearly.Eurycea.trend <- Kendall(Dev.yearly.Eurycea$Year, 
                                    Dev.yearly.Eurycea$Develop.Mean)
summary(Dev.yearly.Eurycea.trend) # p-value = 0.10349

ggplot(Dev.yearly.Eurycea, aes(x = Year, 
                       y = Develop.Mean)) + 
     stat_smooth(method = "lm", 
                 formula = y ~ x, 
                 se = TRUE,
                 alpha = 0.5) +
     xlab("Year") +
     ylab("Mean Developed Land Cover (%)") + 
     ggtitle("Texas Development in Eurycea Stream Catchments over Time")

```

A simple Kendall time-series analysis of the data shows no trend in development throughout Eurycea stream catchments. The test fails to achieve significance (two-sided p-value = 0.10349).

## Question 3: Has development increased over time in new species' habitats?

```{r Eurycea new spp. change over time}

Dev.yearly.new.ts <- ts(Dev.yearly.new$Develop.Mean, 
                    start = c(2001,1,1), end = c(2019,1,1), frequency = 1)
plot(Dev.yearly.new.ts) # Trend is thrown off by the missing values

Dev.yearly.new.trend <- Kendall(Dev.yearly.new$Year, 
                                    Dev.yearly.new$Develop.Mean)
summary(Dev.yearly.new.trend) # p-value = 0.00091197

ggplot(Dev.yearly.new, aes(x = Year, 
                       y = Develop.Mean)) + 
     stat_smooth(method = "lm", 
                 formula = y ~ x, 
                 se = TRUE,
                 alpha = 0.5) +
     xlab("Year") +
     ylab("Mean Developed Land Cover (%)") + 
     ggtitle("Texas Development in Eurycea Stream Catchments over Time")

```

A simple Kendall time-series analysis of the data shows a clear upward trend in development throughout unnamed species' stream catchments. The test achieves significance (two-sided p-value = 0.00091197).

## Question 4: Is development in Texas associated with base flow index?

```{r BFI effect on development}

bfidev.lm <- lm(data = Eurycea, BFI ~ PercentDevelop)
summary(bfidev.lm) # Shows significant negative effect

ggplot(Eurycea, aes(x = PercentDevelop,
                    y = BFI)) + 
     stat_smooth(method = "lm", 
                 formula = y ~ x, 
                 se = TRUE,
                 alpha = 0.5) +
     xlab("Developed Land Cover (%)") +
     ylab("Base Flow Index (%)") + 
     ggtitle("Relationship between Development and Base Flow Index")

```

This simple linear model suggests that base flow index and development cover are negatively correlated. For each 1% increase in development cover within a catchment, BFI decreases by 3%. The model achieves significance (p-value < 2.2e-16), but has an adjusted R-squared of less than 1%.

\newpage

# Summary and Conclusions

Despite the inconvenient format of the dataset, I was able to perform time-series analysis on the percentage of developed land cover within Texas stream catchments from 2001 to 2019. Almost all of my tests achieved significance. From these tests, I can conclude that:

(Question 1) the mean percent of developed land cover within Texas stream catchments has increased from 2001 to 2019;

(Question 2) the mean percent of developed land cover within stream catchments inhabited by all Eurycea species has not changed from 2001 to 2019;

(Question 3) the mean percent of developed land cover within stream catchments inhabited by undescribed Eurycea species has increased from 2001 to 2019;

and (Question 4) as developed land cover within a catchment increases, base flow index within the catchment decreases.

If true, these are sensible and useful results to see. It is broadly understood that Texas is increasingly developing its land as population centers like Austin and San Antonio expand. This development poses a threat to Eurycea habitat. Question 4's linear model evinces this threat: as development increases, base flow index decreases threefold. This is likely because developed areas build wells to extract groundwater for human consumption, reducing the amount of groundwater available to Eurycea in those catchments.

Across Eurycea as a genus, there is no clear trend in development change. This may be because these species occupy different areas within the Edwards-Trinity aquifer system. For example, E. sosorum and E. waterlooensis habitat is centered in Austin. These catchments may already be maximally developed and have little room to change. Other species, such as E. troglodytes, primarily occupy low-development catchments in western Hill Country. These catchments have not yet seen increased development, but may be targeted for development in future. Species like E. neotenes occupy suburban and ex-urban areas, which are at the forefront of development from urban sprawl. This could explain why E. neotenes shows the clearest upward trend in graphs of development change.

However, when subset to undescribed Eurycea species 1, 2, and 3, there is a clear upward trend in development change. This trend is primarily driven by outlier sites for species 3, which have increased from below 30% development to over 35% development. This suggests an increasing rate of development at at least one catchment inhabited by species 3, which may be cause for petitioning USFWS to protect this species.

There are several key weaknesses to this analysis. The adjusted R-squared of Question 4's linear model is minuscule. This suggests that while development may have a relation to base flow index, it is far from the only variable explaining rates of groundwater discharge. Additionally, the time-series data for questions 1-3 is choppy and irregular. This may limit the accuracy of the conclusions drawn.

Overall, this analysis paints a simple picture. Development is increasing in Texas, particularly in catchments occupied by undescribed Eurycea species. This development negatively impacts groundwater discharge, which is crucial for Eurycea habitat. It is clearly imperative to petition USFWS for the protection of these undescribed species.

\newpage

# References

Datri, Crystal (2021). “Petition to the U.S. Fish and Wildlife Service to List the Pedernales River Springs Salamander as a Threatened or Endangered Species Under the Endangered Species Act.”

Devitt, T. J., Wright, A. M., Cannatella, D. C., & Hillis, D. M. (2019). Species delimitation in endangered groundwater salamanders: Implications for aquifer management and biodiversity conservation. Proceedings of the National Academy of Sciences, 116(7), 2624-2633.

Service Completes Initial Reviews on Endangered Species Act Petitions for Four Species | U.S. Fish & Wildlife Service. 2022 Oct 18. FWSgov. [accessed 2023 May 1]. https://fws.gov/press-release/2022-10/service-completes-initial-reviews-endangered-species-act-petitions-four.

Weber M (2023). StreamCatTools: Streamcat Tools. R package version 0.1.1.9002, https://usepa.github.io/StreamCatTools/. 
